# Introduction {#intro}

## How to contribute

(Steps 1. to 5. only need to be done once - to set-up.)

1. Connect your RStudio and GitHub using these instructions, only up to "Create new project" is necessary here (the repository/project already exists):
https://www.datasurg.net/2015/07/13/rstudio-and-github/

2. Get your GitHub account added to the surgicalinformatics organisation on GitHub (ask Riinu/Ewen):
https://github.com/SurgicalInformatics

3. Open the terminal on argonaut, go to your home folder (just `cd` will default to taking you home). If the prompt says `username@argonaut:~$` that means you're home - `~` means user's home.

4. In the terminal, do `git clone https://github.com/SurgicalInformatics/cookbook.git`

5. Look at the Files tab - a new folder called `cookbook` should have appeared at the bottom. Click on it and open the cookbook.Rproj file.

6. Add your thing by editing the appropriate .Rmd file - there's one for each chapter. **click on More - Clean All** (if you don't do this you may be able to compile the book with code that won't work at a subsequent clean build which can be trickier to debug). Use the Build tab to Build your changes into a book.

7. If anyone has pushed since you cloned/last pulled (hopefully they've not been working on the exact same chapter): Make sure you **click on More - Clean All** (as above). Then Pull from the Git tab. This only cleans the output files - html and PDF, it will not touch the changes you've made in the .Rmd file.

8. Then Build Book again - this will include the new changes you pulled as well as your changes.

9. Git tab - commit everything, Push quickly before anyone else does or you'll have to go back to step 7. You can check for new pushed commits here: https://github.com/SurgicalInformatics/cookbook/commits/master Alternatively there's no harm in clicking the Pull button again - it should then say "Already up-to-date".

> Pro tip: instead of clicking on every single file in the Git tab, go to the terminal, `cd cookbook` to go to the project folder if still home, and do `git add -A` which is the same thing. Still need to Commit though!

**10. Have fun!**


## Rules of posting

Rules of how to post here. 

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

## Indexing

### Index

Bold index headings:  
`\index{linear regression@\textbf{linear regression}}` (ticks in .Rmd file are excluded when actually using)

Sub-entries of bold headings:  
`\index{linear regression@\textbf{linear regression}!diagnostics}`

Stand-alone entries:  
`\index{linear regression}`

### Chapter and section references
You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter `\@ref(intro)` (ticks in .Rmd are excluded when actually using). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter `\@ref(methods)`.

### Figure and table references
Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure `\@ref(fig:nice-fig)`. Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table `\@ref(tab:nice-tab)`.

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

### Citations
You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].
