# Snippets

Random useful snippets that do not fit anywhere else.

## Useful RStudio functions

### Clean restart
Cleans current environment and restarts R session (no code can run past this). Indicate objects to leave alone by supplying string to `leave` argument.

```{r message=FALSE, eval=FALSE}
clean_restart = function(leave = NULL){
  rm(list=setdiff(ls(envir = .GlobalEnv), leave), envir = .GlobalEnv)
  invisible(.rs.restartR())
}
```

### Remove duplicated packages between personal and system libraries

This removes any user library packages that are also present in the system library. Use this to minimise loading conflicts. Change `tdrake` in example to your rstudio username.

```{r message=FALSE, eval=FALSE}
library(tidyverse)

tdrake_packages = installed.packages() %>% #change 'tdrake' to your rstudio username
  as_tibble() %>% 
  filter(LibPath %>% str_detect("tdrake")) %>% 
  pull(Package)

system_packages = installed.packages() %>% 
  as_tibble() %>% 
  filter(LibPath %>% str_detect("local")) %>% 
  pull(Package)

to_remove = tdrake_packages[tdrake_packages %in% system_packages]

remove.packages(to_remove)
}
```

### Remove large file accidentally committed to git

`git filter-branch -f --index-filter "git rm -rf --cached --ignore-unmatch filename.rda" HEAD`.


## Memory and efficient RAM usage

1. Restart R often
1. Use `rm(...)` to get rid of big files after using them
1. Save large objects (and compress with "gz") that take a while to create but are needed often (`write_rds(x, "big_file.rds.gz", compress = "gz)`)
1. Run `gc()` (garbage collect)
1. Check memory usage: `pryr::mem_used()`

Most importantly if using big files (even if they aren't loaded in your environment!) do the following:

1. `save.image("temp.rda")`
1. Restart R
1. `load("temp.rda")`
1. Delete "temp.rda" from `Files` pane

## Exporting tables that work in both PDF and Word

The `kable()` function from `library(knitr)` formats tables for all 3 output formats, but it is a bit limited.
`library(kableExtra)` is great for table customisations that work in PDF and HTML.
`kableExtra` functions do not work for Word output, but `library(flextable)` does work for Word (but not for PDF).

We can define our own table printing function that uses kableExtra or flextable based on output type:

```{r message=FALSE}
# This makes table resize or continue over multiple pages in all output types
# PDF powered by kableExtra, Word by flextable
mytable = function(x, caption = "", longtable = FALSE, ...){
  library(kableExtra)
  # if not latex or html then else is Word
  if (knitr::is_latex_output() | knitr::is_html_output()){
    knitr::kable(x, row.names = FALSE, align = c("l", "l", "r", "r", "r", "r", "r", "r", "r"), 
          booktabs = TRUE, caption = caption, longtable = longtable,
          linesep = "", ...) %>%
    kableExtra::kable_styling(latex_options = c("scale_down", "hold_position"))
  }else{
    flextable::flextable(x) %>% 
      flextable::autofit() %>% 
      flextable::width(j = 1, width = 1.5) %>% 
      flextable::height(i = 1, height = 0.5, part = "header")
  }
  
}
library(dplyr)
cars %>% 
  head() %>% 
  mytable()
```

### Warning in kableExtra. Please specify format in kable. kableExtra can customize either HTML or LaTeX outputs.

This warning goes away when you load `library(kableExtra)` in your script, instead of just using `kableExtra::function()`.

What happens when you load `library(kableExtra)` is that it sets that specifies the format for the R session, so that Warning goes away. It does mean that the `render()` function may fail since it knits in the same environment you have, rather than a new one like the Knit button does.



## Creating Reproducible R Examples to Share in the Group (binder, holepunch and docker)

When asking for help with R code having a reproducible example is crucial (some mock data that others can use along with your code to reproduce your error). Often this can be done easily with creation of a small tibble and posting of the code on slack but sometimes it requires more complex data or the error is due to something in the Linux system in which RStudio server is hosted. For example if the `cairo` package for Linux isn't installed then plots don't work. The `holepunch` package helps to reproduce examples like these (not suitable for projects with confidential data).

### Create Basic Reproducible Examples

The three main parts of the reproducible example (reprex) in Surgical Informatics are 1. packages, 2. small dataset and 3. code. Other things like R version and Linux version can be assumed as we all use one of only a few servers.

If you have a small (and confidential) set of data in a tibble or data frame called `my_data` and want it to be easily copied run: `dput(droplevels(my_data))`. This will print out code in the console that can be copy-pasted to reproduce the data frame. Alternatively use the `tibble` or `tribble` functions to create it from scratch (this is preferable for simple datasets). Then copy in the packages and finally the code (ideally the least amount possible to generate the error) and share with the group e.g.:

```{r, error=TRUE, message=FALSE}
library(tidyverse)

# Output generated from dput(droplevels(my_data))
data = structure(list(a = c(1, 2, 3), b = c("a", "b", "c"), c = 10:12), .Names = c("a", 
"b", "c"), row.names = c(NA, -3L), class = c("tbl_df", "tbl", 
"data.frame"))

data %>% 
  mutate(newvar = a /b)
```


### `holepunch` - Complex Reproducible Examples

From your project with data you are happy to make public make sure you are backed up to `git` and `GitHub`. See the relevant chapter on how to do this. Then run the following:

```{r, eval = FALSE}
# Holepunch testing

remotes::install_github("karthik/holepunch")

library(holepunch)
write_compendium_description(package = "Title of my project", 
                             description = "Rough description of project or issue")

write_dockerfile(maintainer = "SurgicalInformatics")

generate_badge()

build_binder()

```

The file will generate some text to copy into the top of a `README.md` file. It will look like:
```{r, eval = FALSE}
<!-- badges: start -->
[![Launch Rstudio Binder](http://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/SurgicalInformatics/<project>/master?urlpath=rstudio)
<!-- badges: end -->
```

Now, whenever somebody clicks on the badge in the `README` on `GitHub` they will be taken to an RStudio server instance with all your files (excluding files listed in `.gitignore`), all the current versions of your package, all the current Linux packages and the current R version. They can then test your code in an near-identical environment to help identify the source of the error, their session will time out after 10 minutes of inactivity or 12 hours since starting and will not save anything so should only be used for bug-testing or quick examples.

As this is a free version of RStudio server there is a limit to what is supported and it shouldn't be used for computationally-intensive processes. 

And, as mentioned: **No confidential data.**

## Working with CHIs

Here are 4 functions for CHIs that could even be put in a small package. 
The Community Health Index (CHI) is a population register, which is used in Scotland for health care purposes. 
The CHI number uniquely identifies a person on the index.

### `chi_dob()` - Extract date of birth from CHI

Note `cutoff_2000`. 
As CHI has only a two digit year, need to decide whether year is 1900s or 2000s. 
I don't think there is a formal way of determining this. 

```{r message=FALSE}
library(dplyr)
chi = c("1009701234", "1811431232", "1304496368")
# These CHIs are not real. 
# The first is invalid, two and three are valid. 

# Cut-off any thing before that number is considered 2000s
# i.e. at cutoff_2000 = 20, "18" is considered 2018, rather than 1918. 
chi_dob = function(.data, cutoff_2000 = 20){
  .data %>% 
    stringr::str_extract(".{6}") %>% 
    lubridate::parse_date_time2("dmy", cutoff_2000 = cutoff_2000) %>% 
    lubridate::as_date() # Make Date object, rather than POSIXct
}

chi_dob(chi)

# From tibble
tibble(chi = chi) %>% 
  mutate(
    dob = chi_dob(chi)
  )
```

### `chi_gender()` - Extract gender from CHI

Ninth digit is odd for men and even for women. 
A test for even is `x modulus 2 == 0`.

```{r}
chi_gender = function(.data){
  .data %>% 
    stringr::str_sub(9, 9) %>% 
    as.numeric() %>% 
    {ifelse(. %% 2 == 0, "Female", "Male")}
}

chi_gender(chi)

# From tibble
tibble(chi = chi) %>% 
  mutate(
    dob = chi_dob(chi),
    gender = chi_gender(chi)
  )
```

### `chi_age()` - Extract age from CHI

Works for a single date or a vector of dates.

```{r message=FALSE}
chi_age = function(.data, ref_date, cutoff_2000 = 20){
  dob = chi_dob(.data, cutoff_2000 = cutoff_2000)
  lubridate::interval(dob, ref_date) %>% 
    as.numeric("years") %>% 
    floor()
}

# Today
chi_age(chi, Sys.time())

# Single date
library(lubridate)
chi_age(chi, dmy("11/09/2018"))

# Vector
dates = dmy("11/09/2018",
            "09/05/2015",
            "10/03/2014")
chi_age(chi, dates)

# From tibble
tibble(chi = chi) %>% 
  mutate(
    dob = chi_dob(chi),
    gender = chi_gender(chi),
    age = chi_age(chi, Sys.time())
  )
```

### `chi_valid()` - Logical test for valid CHI

The final digit of the CHI can be used to test that the number is correct via the modulus 11 algorithm. 

```{r}
chi_valid = function(.data){
  .data %>% 
    stringr::str_split("", simplify = TRUE) %>% 
    .[, -10] %>%              # Working with matrices hence brackets
    apply(1, as.numeric) %>%  # Convert from string
    {seq(10, 2) %*% .} %>%    # Multiply and sum step
    {. %% 11} %>%             # Modulus 11
    {11 - .} %>%              # Substract from 11
    dplyr::near(              # Compare result with 10th digit. 
      {stringr::str_sub(chi, 10) %>% as.numeric()}
    ) %>% 
    as.vector()
}

chi_valid(chi)

# From tibble
tibble(chi = chi) %>% 
  mutate(
    dob = chi_dob(chi),
    gender = chi_gender(chi),
    age = chi_age(chi, Sys.time()),
    chi_valid = chi_valid(chi)
  )
```

## Working with dates

### Difference between two dates

I always forget how to do this neatly. 
I often want days as a numeric, not a lubridate type object. 

```{r}
library(lubridate)
date1 = dmy("12/03/2018", "14/05/2017")
date2 = dmy("11/09/2019", "11/04/2019")

interval(date1, date2) %>% 
  as.numeric("days")
```

### Lags

This is useful for calculating, for instance, the period off medications. Lags are much better than long to wide solutions for this. 

```{r}
library(tidyverse)
library(lubridate)
id = c(2, 2, 2, 2, 3, 5) 
medication = c("aspirin", "aspirin", "aspirin", "tylenol", "lipitor", "advil") 
start.date = c("05/01/2017", "05/30/2017", "07/15/2017", "05/01/2017", "05/06/2017", "05/28/2017")
stop.date = c("05/04/2017", "06/10/2017", "07/27/2017", "05/15/2017", "05/12/2017", "06/13/2017")
df = tibble(id, medication, start.date, stop.date)
df

df %>%
  mutate_at(c("start.date", "stop.date"), lubridate::mdy) %>% # make a date
  arrange(id, medication, start.date) %>% 
  group_by(id, medication) %>% 
  mutate(
    start_date_diff = start.date - lag(start.date),
    medication_period = stop.date-start.date
  )
```

### Pulling out "change in status" data

If you have a number of episodes per patient, each with a status and a time, then you need to do this as a starting point for CPH analysis. 

### Guess the Format of Dates

This function helps guess the date format from a dataset you have received. Add in new formats (as per lubridate) if they don't match. Once you work out what format it is then extract date with `mdy()` or `dmy()` etc. based on format.

```{r, eval = FALSE, echo = TRUE}
library(tidyverse)
library(lubridate)
lakers = lakers %>%
  mutate(mm_date = paste0(str_sub(date, start =5, end = 6), "/",
                          str_sub(date, start = 7, end = 8), "/",
                          str_sub(date, start = 1, end = 4)),
         dd_date = paste0(str_sub(date, start =7, end = 8), "/",
                          str_sub(date, start = 5, end = 6), "/",
                          str_sub(date, start = 1, end = 4)))

guess_formats(lakers$mm_date, c("dmy", "mdy"), print_matches = TRUE)
guess_formats(lakers$dd_date, c("dmy", "mdy"), print_matches = TRUE)
```


#### Example data

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(lubridate)
library(finalfit)
mydata = tibble(
  id = c(1,1,1,1,2,2,2,2,3,3,3,3,4,4,4,4,5,5,5,5),
  status = c(0,0,0,1,0,0,1,1,0,0,0,0,0,1,1,1,0,0,1,1),
  group = c(rep(0, 8), rep(1, 12)) %>% factor(),
  opdate = rep("2010/01/01", 20) %>% ymd(),
  status_date = c(
    "2010/02/01", "2010/03/01", "2010/04/01", "2010/05/01",
    "2010/02/02", "2010/03/02", "2010/04/02", "2010/05/02",
    "2010/02/03", "2010/03/03", "2010/04/03", "2010/05/03",
    "2010/02/04", "2010/03/04", "2010/04/04", "2010/05/04",
    "2010/02/05", "2010/03/05", "2010/04/05", "2010/05/05"
  ) %>% ymd()
)
mydata
```

#### Compute time from op date to current review
... if necessary

```{r}
mydata = mydata %>% 
  arrange(id, status_date) %>% 
  mutate(
    time = interval(opdate, status_date) %>% as.numeric("days")
  )
mydata
```

#### Pull out "change of status"
```{r}
mydata = mydata %>% 
  group_by(id) %>% 
  mutate(
    status_change = status - lag(status) == 1,                          # Mark TRUE if goes from 0 to 1
    status_nochange = sum(status) == 0,                                 # Mark if no change from 0
    status_nochange_keep = !duplicated(status_nochange, fromLast= TRUE) # Mark most recent "no change" episode
  ) %>% 
  filter(status_change | (status_nochange & status_nochange_keep)) %>%  # Filter out other episodes
  select(-c(status_change, status_nochange, status_nochange_keep))      # Remove columns not needed
mydata
```

#### Run CPH
```{r}
mydata %>% 
  finalfit("Surv(time, status)", "group")
```

### Isoweek and epiweek

This is frequently painful, when working with weeks in a year to aggregate data, how to best plot and how to deal with multiple years. 

Easiest solution I have found (EMH) is just to work with the first date of each isoweek. Make a look up table and join. 

```{r}
# Generate random set of dates for demonstration purposes
## This is not required when using normally
library(tidyverse)
library(lubridate)
library(finalfit)
colon_s = colon_s %>% 
  mutate(
    date = seq(ymd(20200101), ymd(20220601), by="day") %>% 
      sample(929, replace = TRUE)
  )

# Make isoweek start date look-up table
## Make sure includes entire range of your dates of interest. 
isoweek_lookup = tibble(date = seq(ymd(20200101), ymd(20220601), by = 1)) %>% 
  mutate(isoweek = lubridate::isoweek(date),
         year = lubridate::year(date))

isoweek_lookup = isoweek_lookup %>% 
  left_join(isoweek_lookup %>% 
              group_by(isoweek, year) %>% 
              slice(1) %>% 
              select(isoweek_date = date, isoweek, year)
  )

# Add isoweek_date (first date of a particular isoweek) and isoweek to analysis dataset. 
colon_s %>% 
  left_join(isoweek_lookup, by = c("date" = "date")) %>% 
  select(date, isoweek, isoweek_date) %>% 
  slice(1:6)
```



## Labelling tables and figures in markdown chunks

### Plots

I have made a plot and it looks nice (See Figure `\ref{fig:a-nice-plot}`)

```` markdown
`r ''````{r fig.cap="A nice plot.\\label{fig:a-nice-plot}}
ggplot(mydata, aes(x = x)) +
  geom_histogram()
```
````

### Tables

I have made a table and it looks nice (See Table `\ref{tab:a-nice-table}`)

```` markdown
`r ''````{r}
library(knitr)
kable(mtcars[1:3,1:4], caption="A nice table.\\label{tab:a-nice-table}")
```
````



## Join tables and overwrite values (that are often missing)

From here: https://stackoverflow.com/questions/35637135/left-join-two-data-frames-and-overwrite

```{r}
library(dplyr)
.keys <- c("key1", "key2")

.base_table <- tribble(
    ~key1, ~key2, ~val1, ~val2,
    "A", "a", 0, 0,
    "A", "b", 0, 1,
    "B", "a", 1, 0,
    "B", "b", 1, 1)

.join_table <- tribble(
    ~key1, ~key2, ~val2,
    "A", "b", 100,
    "B", "a", 111)

# This works
df_result <- .base_table %>%
    # Pull off rows from base table that match the join table
    semi_join(.join_table, .keys) %>%
    # Drop cols from base table that are in join table, except for the key columns
    select(-matches(setdiff(names(.join_table), .keys))) %>%
    # Left join on the join table columns
    left_join(.join_table, .keys) %>%
    # Remove the matching rows from the base table, and 
    # bind on the newly joined result from above.
    bind_rows(.base_table %>% anti_join(.join_table, .keys))

df_result %>%
    print()
```

## Read in multiple files

As of July 2021, library(readr) is happy to read in multiple files into the same tibble without you having to use `purrr::map_dfr()`.

```{r, eval = FALSE}
library(tidyverse)

exampledata = read_csv(fs::dir_ls(path = "data_folder", glob = "*.csv"))
```

It's also happy to automatically uncompress files ending with .zip, .gz or similar.

## Averaging multiple AUCs (say from imputed datasets)

Final method here probably makes most sense: `

```{r eval=FALSE}
auc = c(0.5, 0.6, 0.7, 0.8, 0.9, 0.99999)
mean(auc)                       # Mean 
median(auc)                     # Median
prod(auc)^(1/length(auc))       # Geometric mean
exp(mean(log(auc)))             # Geometric mean again
log(auc)                        # Logs
log(auc / (1-auc) )             # Put values on scale from 0 (0.5) to Inf (1)
exp(mean(log(auc/(1-auc)))) /   # Mean on the above scale. 
  (1 + exp(mean(log(auc/(1-auc)))))
```

```{r}
auc = c(0.5, 0.6, 0.7, 0.8, 0.9, 0.99999)
mean(auc)                       # Mean 
median(auc)                     # Median
prod(auc)^(1/length(auc))       # Geometric mean
exp(mean(log(auc)))             # Geometric mean again
log(auc)                        # Logs
log(auc / (1-auc) )             # Put values on scale from 0 (0.5) to Inf (1)
exp(mean(log(auc/(1-auc)))) /   # Mean on the above scale. 
  (1 + exp(mean(log(auc/(1-auc)))))
```
