# Snippets

Random useful snippets that do not fit anywhere else. 

## Working with CHIs

Here are 4 functions for CHIs that could even be put in a small package. 
The Community Health Index (CHI) is a population register, which is used in Scotland for health care purposes. 
The CHI number uniquely identifies a person on the index.

### `chi_dob()` - Extract date of birth from CHI

Note `cutoff_2000`. 
As CHI has only a two digit year, need to decide whether year is 1900s or 2000s. 
I don't think there is a formal way of determining this. 

```{r message=FALSE}
library(dplyr)
chi = c("1009701234", "1811431232", "1304496368")
# These CHIs are not real. 
# The first is invalid, two and three are valid. 

# Cut-off any thing before that number is considered 2000s
# i.e. at cutoff_2000 = 20, "18" is considered 2018, rather than 1918. 
chi_dob = function(.data, cutoff_2000 = 20){
  .data %>% 
    stringr::str_extract(".{6}") %>% 
    lubridate::parse_date_time2("dmy", cutoff_2000 = cutoff_2000) %>% 
    lubridate::as_date() # Make Date object, rather than POSIXct
}

chi_dob(chi)

# From tibble
tibble(chi = chi) %>% 
  mutate(
    dob = chi_dob(chi)
  )
```

### `chi_gender()` - Extract gender from CHI

Ninth digit is odd for men and even for women. 
A test for even is `x modulus 2 == 0`.

```{r}
chi_gender = function(.data){
  .data %>% 
    stringr::str_sub(9, 9) %>% 
    as.numeric() %>% 
    {ifelse(. %% 2 == 0, "Female", "Male")}
}

chi_gender(chi)

# From tibble
tibble(chi = chi) %>% 
  mutate(
    dob = chi_dob(chi),
    gender = chi_gender(chi)
  )
```

### `chi_age()` - Extract age from CHI

Works for a single date or a vector of dates.

```{r message=FALSE}
chi_age = function(.data, ref_date, cutoff_2000 = 20){
  dob = chi_dob(.data, cutoff_2000 = cutoff_2000)
  lubridate::interval(dob, ref_date) %>% 
    as.numeric("years") %>% 
    floor()
}

# Today
chi_age(chi, Sys.time())

# Single date
library(lubridate)
chi_age(chi, dmy("11/09/2018"))

# Vector
dates = dmy("11/09/2018",
            "09/05/2015",
            "10/03/2014")
chi_age(chi, dates)

# From tibble
tibble(chi = chi) %>% 
  mutate(
    dob = chi_dob(chi),
    gender = chi_gender(chi),
    age = chi_age(chi, Sys.time())
  )
```

### `chi_valid()` - Logical test for valid CHI

The final digit of the CHI can be used to test that the number is correct via the modulus 11 algorithm. 

```{r}
chi_valid = function(.data){
  .data %>% 
    stringr::str_split("", simplify = TRUE) %>% 
    .[, -10] %>%              # Working with matrices hence brackets
    apply(1, as.numeric) %>%  # Convert from string
    {seq(10, 2) %*% .} %>%    # Multiply and sum step
    {. %% 11} %>%             # Modulus 11
    {11 - .} %>%              # Substract from 11
    dplyr::near(              # Compare result with 10th digit. 
      {stringr::str_sub(chi, 10) %>% as.numeric()}
    ) %>% 
    as.vector()
}

chi_valid(chi)

# From tibble
tibble(chi = chi) %>% 
  mutate(
    dob = chi_dob(chi),
    gender = chi_gender(chi),
    age = chi_age(chi, Sys.time()),
    chi_valid = chi_valid(chi)
  )
```

## Working with dates

### Difference between two dates

I always forget how to do this neatly. 
I often want days as a numeric, not a lubridate type object. 

```{r}
library(lubridate)
date1 = dmy("12/03/2018", "14/05/2017")
date2 = dmy("11/09/2019", "11/04/2019")

interval(date1, date2) %>% 
  as.numeric("days")
```

### Lags

This is useful for calculating, for instance, the period off medications. Lags are much better than long to wide solutions for this. 

```{r}
library(tidyverse)
library(lubridate)
id = c(2, 2, 2, 2, 3, 5) 
medication = c("aspirin", "aspirin", "aspirin", "tylenol", "lipitor", "advil") 
start.date = c("05/01/2017", "05/30/2017", "07/15/2017", "05/01/2017", "05/06/2017", "05/28/2017")
stop.date = c("05/04/2017", "06/10/2017", "07/27/2017", "05/15/2017", "05/12/2017", "06/13/2017")
df = tibble(id, medication, start.date, stop.date)
df

df %>%
  mutate_at(c("start.date", "stop.date"), lubridate::mdy) %>% # make a date
  arrange(id, medication, start.date) %>% 
  group_by(id, medication) %>% 
  mutate(
    start_date_diff = start.date - lag(start.date),
    medication_period = stop.date-start.date
  )
```

### Pulling out "change in status" data

If you have a number of episodes per patient, each with a status and a time, then you need to do this as a starting point for CPH analysis. 

#### Example data

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(lubridate)
library(finalfit)
mydata = tibble(
  id = c(1,1,1,1,2,2,2,2,3,3,3,3,4,4,4,4,5,5,5,5),
  status = c(0,0,0,1,0,0,1,1,0,0,0,0,0,1,1,1,0,0,1,1),
  group = c(rep(0, 8), rep(1, 12)) %>% factor(),
  opdate = rep("2010/01/01", 20) %>% ymd(),
  status_date = c(
    "2010/02/01", "2010/03/01", "2010/04/01", "2010/05/01",
    "2010/02/02", "2010/03/02", "2010/04/02", "2010/05/02",
    "2010/02/03", "2010/03/03", "2010/04/03", "2010/05/03",
    "2010/02/04", "2010/03/04", "2010/04/04", "2010/05/04",
    "2010/02/05", "2010/03/05", "2010/04/05", "2010/05/05"
  ) %>% ymd()
)
mydata
```

#### Compute time from op date to current review
... if necessary

```{r}
mydata = mydata %>% 
  arrange(id, status_date) %>% 
  mutate(
    time = interval(opdate, status_date) %>% as.numeric("days")
  )
mydata
```

#### Pull out "change of status"
```{r}
mydata = mydata %>% 
  group_by(id) %>% 
  mutate(
    status_change = status - lag(status) == 1,                          # Mark TRUE if goes from 0 to 1
    status_nochange = sum(status) == 0,                                 # Mark if no change from 0
    status_nochange_keep = !duplicated(status_nochange, fromLast= TRUE) # Mark most recent "no change" episode
  ) %>% 
  filter(status_change | (status_nochange & status_nochange_keep)) %>%  # Filter out other episodes
  select(-c(status_change, status_nochange, status_nochange_keep))      # Remove columns not needed
mydata
```

#### Run CPH
```{r}
mydata %>% 
  finalfit("Surv(time, status)", "group")
```
