# Data manipulation

## Collapse multiple "no" and "yes" options

Common to have to do this in GlobalSurg projects:

```{r message=FALSE}
library(dplyr)
mydata = tibble(
  ssi.factor = c("No", "Yes, no treatment/wound opened only (CD 1)",    
                 "Yes, antibiotics only (CD 2)", "Yes, return to operating theatre (CD 3)", 
                 "Yes, requiring critical care admission (CD 4)", 
                 "Yes, resulting in death (CD 5)",
                 "Unknown") %>%
    factor(),
  mri.factor = c("No, not available", "No, not indicated", 
                 "No, indicated and facilities available, but patient not able to pay",
                 "Yes", "Unknown", "Unknown", "Unknown") %>% 
    factor()
)

# Two functions make this work
fct_collapse_yn = function(.f){
  .f %>% 
    forcats::fct_relabel(~ gsub("^No.*", "No", .)) %>% 
    forcats::fct_relabel(~ gsub("^Yes.*", "Yes", .))
}

is.yn = function(.data){
  .f = is.factor(.data)
  .yn = .data %>% 
    levels() %>% 
    grepl("(No.+)|(Yes.+)", .) %>% 
    any()
  all(.f, .yn)
}

# Raw variable
mydata %>% 
  count(ssi.factor)

# Collapse to _yn version
mydata %>% 
  mutate(across(where(is.yn), fct_collapse_yn, .names = "{col}_yn")) %>% 
  count(ssi.factor_yn)

```


## Filter NA: Dropping rows where all specified variables are NA


I want to keep rows that have a value for `important_a` and/or `important_b` (so rows 1, 3, 4).
I don't care whether `whatever_c` is empty or not, but I do want to keep it.

```{r, message=FALSE}
library(tidyverse)

mydata  = tibble(important_a = c("Value", NA, "Value", NA, NA),
                 important_b = c(NA, NA, "Value", "Value", NA),
                 whatever_c  = c(NA, "Value", NA, NA, NA))

mydata
```



Functions for missing values that are very useful, but don't do what I want are:

(1) This keeps complete cases based on all columns:
```{r}
mydata %>% 
  drop_na()
```

(Returns 0 as we don't have rows where all 3 columns have a value).

(2) This keeps complete cases based on specified columns:

```{r}
mydata %>% 
  drop_na(important_a, important_b)
```
This only keeps the row where both a and b have a value.


(3) This keeps rows that have a value in any column:

```{r}
mydata %>% 
  filter_all(any_vars(! is.na(.)))
```


The third example is better achieved using the janitor package:

```{r, message = FALSE}
mydata %>% 
  janitor::remove_empty()
```

Now, (3) is pretty close, but still, I'm not interested in row 2 - where both a and b are empty but c has a value (which is why it's kept).

(4) Simple solution

A quick solution is to use `! is.na()` for each variable inside a `filter()`:

```{r}
mydata %>% 
  filter(! is.na(important_a) | ! is.na(important_b))
```

And this is definitely what I do when I only have a couple of these variables. But if you have tens, then the filtering logic becomes horrendously long and it's easy to miss one out/make a mistake.

(5) Powerful solution:

A scalable solution is to use `filter_at()` with `vars()` with a select helper (e.g., `starts with()`), and then the `any_vars(! is.na(.))` that was introduced in (3).

```{r}
mydata %>% 
  filter_at(vars(starts_with("important_")), any_vars(! is.na(.)))
```







